% ============================================================
%  05.tex — Computer-algebra foundations: Symbolica and Spenso
% ============================================================

% ==============================================================
\chapter{Computer-Algebra Foundations: Symbolica and Spenso}
\label{ch:cas}
% ==============================================================

Before describing how the toolkit turns a declared model into a Feynman rule (Chapter~\ref{ch:internals}), it is worth spending a chapter on the two libraries the whole project is built on. Every symbolic object the toolkit manipulates --- a coupling, a momentum, a $\gamma$ matrix, an entire vertex --- is ultimately an object provided by \textsc{Symbolica}, and every \emph{index} carried by those objects is given meaning by \textsc{Spenso}. The clean division of labour between the two is precisely what makes the pipeline of the next chapter feasible: \textbf{Symbolica manipulates expressions, Spenso tells Symbolica what the indices inside those expressions mean, and a small companion layer, idenso, knows the physics-specific identities those typed indices satisfy.} This chapter explains what each library is, how it works, and why it was chosen.

% --------------------------------------------------------------
\section{Symbolica: a modern computer-algebra core}
\label{sec:cas-symbolica}
% --------------------------------------------------------------

\textsc{Symbolica}~\cite{symbolica} is a high-performance computer-algebra system (CAS): a Rust engine exposed through native Python, Rust and C++ bindings, developed by Ben Ruijl and used in high-energy-physics research at, among others, CERN. It grew directly out of the practical needs of collider physics, where symbolic computations routinely generate expressions with billions of terms that can occupy terabytes of memory, and where the established tools (the \textsc{Form} lineage, and general-purpose systems such as \textsc{Mathematica}) are either hard to maintain or not built for that scale~\cite{symbolica,ruijl_symbolica_talk}. Symbolica is source-available on GitHub and free for hobbyist and non-commercial single-core use, with a license required for professional or institutional work.

For the purposes of this thesis, four properties of Symbolica are directly responsible for design choices made throughout the toolkit.

\subsection*{Native embedding, no glue code}

Symbolica is linked directly into the running Python process; it is not an external program driven through temporary files or a text pipe, as one would drive \textsc{Form} or a \textsc{Mathematica} kernel. There is therefore no string round-tripping and no serialisation boundary: a Symbolica \py{Expression} is a first-class Python value that can be stored on a dataclass field (as the toolkit does with \py{InteractionTerm.coupling}), compared, hashed, put in a set, and passed between functions like any other object. This is what allows the toolkit to treat ``a piece of a Lagrangian'' and ``a partially built Feynman rule'' as ordinary Python data structures rather than as opaque strings.

\subsection*{Expressions, symbols, and pattern matching}

The user-facing surface of Symbolica is small and regular. Symbols are created with \py{S("name")} (and mathematical expressions can also be parsed from strings with \py{E("...")}); ordinary Python arithmetic operators then build up expressions:

\begin{minted}{python}
from symbolica import S, E, Expression
x, y, f = S("x", "y", "f")
e = f((1 + x)**2)          # a function application, f of (1+x)^2
\end{minted}

The single most important capability for this project is \emph{pattern matching with wildcards}. Any symbol whose name ends in an underscore is treated as a wildcard in \py{match}/\py{replace} operations, so algebraic rewriting reads almost like a regular expression for mathematics:

\begin{minted}{python}
x_, n_ = S("x_", "n_")
e = f((1 + x)**2).replace(f(x_**n_), f(x_)**n_)
print(e)   # ->  f(1 + x)**2
\end{minted}

The toolkit leans on this pervasively. As we will see in Chapter~\ref{ch:internals}, the central step of the contraction engine --- attaching an external leg's index label to an open index of a coupling --- is literally one \py{coupling.replace(open_label, target_label)} call, and the user-facing helper \py{pattern_coefficient} (Section~\ref{sec:feynpy-expr-manip}) is a thin wrapper around the same wildcard machinery. Standard CAS operations are available on the same objects: \py{.expand()}, \py{.collect()}, \py{.derivative()} (with the chain and product rules applied automatically), \py{.factor()}, \py{.coefficient()}, series expansion, and even \py{solve_linear_system}.

\subsection*{Polynomial algebra: the source of the speed}

Underneath the expression layer sits a world-class implementation of multivariate polynomial arithmetic --- greatest common divisors, factorisation, and interpolation over the rationals~\cite{symbolica}. This is the ``beating heart'' of Symbolica's performance, and the gap over general-purpose systems is not marginal: on Symbolica's own multivariate-GCD benchmark, a computation that takes about $89\,\mathrm{s}$ in \textsc{Mathematica} and over an hour in \textsc{SymPy} completes in roughly $4\,\mathrm{s}$~\cite{symbolica}. This matters here because the contraction engine of Chapter~\ref{ch:internals} generates one term per permutation of the external legs and then relies on Symbolica's \py{.expand()}/\py{.collect()} to bring the resulting, often large, sums back to a compact polynomial-in-momenta form; the cost of that repeated normalisation is exactly where a fast polynomial core pays off.

\subsection*{Tensor and graph canonicalisation}

Beyond scalar algebra, Symbolica can put \emph{tensor} expressions into a canonical form. Given a product of tensors with contracted (dummy) indices, \py{Expression.canonize_tensors} relabels the dummies deterministically and orders the factors, taking into account symmetry metadata attached to each tensor head through the \py{is_symmetric}/\py{is_antisymmetric} flags on its symbol. Internally this rests on a graph-canonicalisation algorithm for (cyclic-)symmetric tensor networks~\cite{ruijl_symbolica_talk}. Two expressions that are equal up to a renaming of summed indices and a reordering of commuting factors are thereby reduced to the same normal form --- which is exactly what is needed to compare a derived vertex against a reference, or to check that a gauge variation vanishes. The toolkit's own canonicalisation layer (Section~\ref{sec:internals-postprocessing}) is built directly on top of this primitive.

\begin{tcolorbox}[infobox, title={Why a fast CAS matters here}]
The toolkit does not perform a single, closed-form derivation; it \emph{expands} covariant derivatives and field strengths into many elementary monomials and then sums over all Wick contractions of each. Both steps multiply the number of terms. A CAS that treats large expressions as ordinary in-process objects, rewrites them by pattern matching, and normalises them through a fast polynomial core is therefore not a convenience but a prerequisite.
\end{tcolorbox}

% --------------------------------------------------------------
\section{Spenso: typed tensor indices and tensor networks}
\label{sec:cas-spenso}
% --------------------------------------------------------------

Symbolica by itself has no notion of a tensor index. To Symbolica, a symbol such as \py{G(mu, a)} is merely a function applied to two opaque arguments; nothing records that \py{mu} lives in four-dimensional Minkowski space, that \py{a} is an $\mathrm{SU}(3)$ adjoint index, or that writing the same label \py{a} on two different factors is meant to denote a sum over the adjoint representation. \textsc{Spenso}~\cite{spenso} supplies exactly this missing layer. It is a Rust tensor library --- ``a sparse and dense tensor library with support for automatic matching indices and tensor networks'' --- designed for high-energy-physics calculations and integrated with Symbolica, and it is used from Python through the \py{symbolica.community.spenso} module.

\subsection*{Slots and representations}

The central concept is the \py{Slot}: a typed index, i.e.\ a pair \py{(Representation, abstract_index)}. A \py{Representation} fixes both the \emph{dimension} of an index and its \emph{transformation law}. The toolkit only ever uses six of them (Table~\ref{tab:spenso-representations}), each constructed once and then reused for every slot of that kind.

\begin{table}[htbp]
\centering
\begin{tabular}{@{}llp{6.6cm}@{}}
\toprule
\textbf{Constructor} & \textbf{Self-dual?} & \textbf{Used for} \\
\midrule
\texttt{Representation.mink(4)}  & yes & Lorentz vector indices ($\mu,\nu,\dots$) \\
\texttt{Representation.bis(4)}   & yes & Dirac bispinor indices on fermion fields \\
\texttt{Representation.cof(3)}   & no  & $\mathrm{SU}(3)$ colour fundamental (quark) index \\
\texttt{Representation.coad(8)}  & yes & $\mathrm{SU}(3)$ colour adjoint (gluon / generator / $f^{abc}$) index \\
\texttt{Representation.cof(2)}   & no  & $\mathrm{SU}(2)_L$ weak-isospin fundamental (doublet) index \\
\texttt{Representation.coad(3)}  & yes & $\mathrm{SU}(2)_L$ weak-isospin adjoint (triplet) index \\
\bottomrule
\end{tabular}
\caption{The Spenso representations used by the toolkit. Each is wrapped by one of the toolkit's own \texttt{IndexType} constants of Section~\ref{sec:feynpy-indices} (\texttt{LORENTZ\_INDEX}, \texttt{SPINOR\_INDEX}, \texttt{COLOR\_FUND\_INDEX}, \texttt{COLOR\_ADJ\_INDEX}, \texttt{WEAK\_FUND\_INDEX}, \texttt{WEAK\_ADJ\_INDEX}).}
\label{tab:spenso-representations}
\end{table}

The \emph{self-dual} column is not an implementation detail: it is the tensor-index counterpart of the representation theory of Chapter~\ref{ch:theory}. A colour \emph{fundamental} index (carried by a quark $q^i$) and its dual, the \emph{antifundamental} index (carried by an antiquark $\bar q_i$), are genuinely different representations. Consequently \py{cof(3)} is \emph{dualizable}: Spenso keeps track of which of the two a given slot is, and only allows a fundamental slot to contract with an antifundamental one, exactly as one must distinguish upper from lower colour indices by hand. A Lorentz, Dirac or colour-\emph{adjoint} index, on the other hand, is equivalent to its own dual --- there is a metric (or Killing form) that raises and lowers it --- so \py{mink}, \py{bis} and \py{coad} are self-dual and any two matching slots of that kind may contract directly. Spenso enforces these rules automatically, which is why the toolkit's own bookkeeping never has to reason about representations directly: it only needs to track index \emph{labels}, and let Spenso decide what is contractible.

\subsection*{Building typed tensors}

A physical tensor is built by calling a \py{TensorName} --- a tagged head recognised by Spenso's HEP tensor library, such as \py{TensorName.gamma()}, \py{TensorName.t()}, or \py{TensorName.f()} --- on a list of typed slots, and then converting the result to an ordinary Symbolica expression with \py{.to_expression()} so it can be multiplied into a Lagrangian term like any other factor. The toolkit collects these constructors in one module (\texttt{symbolic.spenso\_structures}); two representative examples are:

\begin{minted}{python}
def gamma_matrix(left_spinor, right_spinor, lorentz):
    """gamma^mu as a Spenso tensor: bispinor x bispinor x Minkowski."""
    return TensorName.gamma()(
        bispinor_index(left_spinor),   # Representation.bis(4)(left_spinor)
        bispinor_index(right_spinor),  # Representation.bis(4)(right_spinor)
        lorentz_index(lorentz),        # Representation.mink(4)(lorentz)
    ).to_expression()

def gauge_generator(adj_index, fund_left, fund_right):
    """t^a_{ij}: adjoint x fundamental x fundamental."""
    return TensorName.t()(
        color_adj_index(adj_index),    # Representation.coad(8)(adj_index)
        color_fund_index(fund_left),   # Representation.cof(3)(fund_left)
        color_fund_index(fund_right),  # Representation.cof(3)(fund_right)
    ).to_expression()
\end{minted}

\noindent The structure constant $f^{abc}$, the weak-sector generators and structure constants, $\gamma_5$, $\sigma^{\mu\nu}$ and the metrics follow the identical pattern. Because every slot already carries its \py{Representation}, it is Spenso --- not the toolkit --- that knows the two \py{a}-labelled adjoint slots on \py{t(a,i,j)} and \py{G(mu,a)} are contractible.

\subsection*{Tensor networks and automatic index matching}

The feature that gives Spenso its name is the \py{TensorNetwork}. A product of typed tensors is a network whose edges are the shared index labels; Spenso can \emph{parse such a network directly from a Symbolica expression}, match indices automatically, and contract the network --- either numerically, using stored components, or symbolically. Data can be stored densely (\py{DenseTensor}) or sparsely (\py{SparseTensor}), and a tensor with no data attached behaves as a purely symbolic object. The toolkit uses this both ways: most of the pipeline treats tensors symbolically, but for the custom invariants it defines (the $\mathrm{SU}(2)$ doublet $\varepsilon_{ij}$, the Minkowski and colour Levi-Civita symbols, the totally symmetric $d^{abc}$, and the charge-conjugation matrix) it also registers explicit numeric components, so that a network mixing standard HEP tensors with these invariants can be evaluated all the way down to a scalar.

\begin{tcolorbox}[notebox, title={Custom invariants}]
Spenso ships components for the standard HEP tensors ($\gamma$, $\sigma$, generators, metrics) in \py{TensorLibrary.hep_lib()}. The project extends this library with numeric components for its own antisymmetric invariants so they can participate in tensor-network execution. Their \emph{symmetry}, however, is not attached to the live Spenso head; it is registered separately at the canonicalisation layer (Section~\ref{sec:internals-postprocessing}), which keeps the Symbolica namespace stable across the pipeline.
\end{tcolorbox}

% --------------------------------------------------------------
\section{idenso: physics-aware simplification}
\label{sec:cas-idenso}
% --------------------------------------------------------------

The third piece is \textbf{idenso}, a companion to Spenso that supplies physics-aware simplification passes over typed tensor networks. Three are used by the toolkit:

\begin{itemize}
    \item \py{simplify_metrics} contracts chains of Kronecker deltas and metric tensors, e.g.\ $g^{\mu\nu} g_{\nu\rho} \to \delta^\mu{}_\rho$, and absorbs metrics contracted against other tensors;
    \item \py{simplify_color} applies $\mathrm{SU}(N)$ colour identities --- Fierz-type collapses of networks of generators $t^a$ and structure constants $f^{abc}$;
    \item \py{simplify_gamma} performs the Clifford (Dirac) algebra on chains of $\gamma$ matrices.
\end{itemize}

The toolkit composes the three into one pass, \py{simplify_invariants}, run in a deliberately fixed order --- metrics, then colour, then gamma, then metrics again --- because each pass can expose fresh metric pairs left behind by the previous one. Each idenso step is a no-op on any subexpression it does not recognise, so the composed pass is safe to call on any vertex factor. This pass, together with the tensor-canonicalisation step built on Symbolica's \py{canonize_tensors} (Section~\ref{sec:internals-postprocessing}), is what turns the raw permutation sum produced by the contraction engine into the compact vertex that \py{feynman_rule(...)} finally prints.

% --------------------------------------------------------------
\section{The bridge: \texttt{IndexType}}
\label{sec:cas-indextype}
% --------------------------------------------------------------

The toolkit ties the three libraries together with one small object, \py{IndexType}, discussed from the user's point of view in Section~\ref{sec:feynpy-indices}. Each \py{IndexType} pairs a single Spenso \py{Representation} with a plain string \emph{kind} (\py{"lorentz"}, \py{"spinor"}, \py{"color_fund"}, \dots) and a default label prefix. The \py{Representation} is what Spenso needs to decide contractibility; the \py{kind} string is a cheap key that the toolkit's own compiler uses to group and match indices without repeatedly inspecting the representation object. Every index that appears anywhere in the pipeline --- on a field, on a derivative, on a coupling --- is tagged with an \py{IndexType}, and it is this tag that lets the layers of Chapter~\ref{ch:internals} reason about indices uniformly, from the declarative front end all the way down to the contraction engine.

In summary, the division of responsibility is sharp and deliberate:
\begin{center}
\begin{tabular}{@{}lp{9.2cm}@{}}
\toprule
\textbf{Layer} & \textbf{Responsibility} \\
\midrule
\textsc{Symbolica} & Represent and manipulate symbolic expressions; pattern matching; polynomial normalisation; tensor/graph canonicalisation. \\
\textsc{Spenso}    & Give indices a representation and dimension; decide which indices contract; build and evaluate tensor networks. \\
idenso             & Apply physics identities (metrics, $\mathrm{SU}(N)$ colour, Dirac algebra) to typed tensor networks. \\
\py{IndexType}     & Bridge the above to the toolkit by pairing one representation with a stable ``kind'' key. \\
\bottomrule
\end{tabular}
\end{center}

With these foundations in place, the next chapter can describe the pipeline that consumes them: how a declared \py{Model} is progressively rewritten into elementary interaction terms and then contracted, through Symbolica and Spenso, into a Feynman rule.
