% ============================================================
%  06.tex — Implementation: from Model to Feynman Rule
% ============================================================

% ==============================================================
\chapter{Implementation: From \texttt{Model} to Feynman Rule}
\label{ch:internals}
% ==============================================================
\cleanup{from here started}
Chapter~\ref{ch:feynpy} served as a user guide, whilst Chapter~\ref{ch:cas} outlined the libraries on which FEYNPY is based. Now that we know how to use the toolkit and understand its language which, although it is Python, may not be familiar to everyone, we can move on to understanding how it works internally. How do we get from the declarative Python objects that the user writes to the Symbolica expression that emerges as a Feynman rule? 

We physicists, in understanding the origin and physical meaning of a theory, make use of many conventions and simplifications (which, moreover, differ from physicist to physicist), and the code must recognise these conventions. When considering the Standard Model, for example, nobody would want to write every index, every generator and every tensor component explicitly. This would be a pointless and tedious task, and if not done with care and focus, it could even lead to unintentional errors. The toolkit is therefore intended to take a compact input and expand it according to the conventions and declarations encoded in the model.

Internally, the code will have to carry out a great deal of bookkeeping. In other words, it must be capable of expanding terms such as the covariant derivative according to the correct rules dictated by the symmetry groups that each Model possesses. It must be able to keep track of the order of the fields and their different indices. To do this a sequence of small rewriting steps, each of which makes the expression slightly more explicit is applied. One can visualise the structure as an hourglass: the user-facing interface allows several compact ways of declaring a Lagrangian, but these different inputs are progressively rewritten until they meet at the central object, the InteractionTerm. From this point on, the code can perform Wick contractions, canonicalisation and simplification of the resulting expression.

We want to proceed having an example in mind. Let's consider the quark kinetic term of QCD,
\begin{equation}
    i\,\bar q\,\gamma^\mu D_\mu q \;\supset\; g_s\,\bar q_{i}\,\gamma^\mu\,t^a_{ij}\,q_{j}\,G^a_\mu ,
    \label{eq:qcd-current}
\end{equation}
is, by the end of the pipeline, turned into the quark--gluon vertex $-i g_s\,\gamma^\mu\,t^a_{ij}$. The full gauge-fixed QCD Lagrangian,
\begin{equation}
\mathcal{L}_{\mathrm{QCD}}
= i\,\bar q\,\gamma^\mu D_\mu q
\;-\;\tfrac14 F^a_{\mu\nu} F^{a\,\mu\nu}
\;-\;\tfrac{1}{2\xi}\!\left(\partial^\mu A^a_\mu\right)^2
\;-\;\bar c^{\,a}\,\partial^\mu (D_\mu c)^a ,
\label{eq:qcd-full}
\end{equation}
is a good stress test because it exercises every branch of the compiler at once: a covariant matter kinetic term, a field strength, a gauge-fixing term and a ghost term.

% --------------------------------------------------------------
\section{Overview of the pipeline}
\label{sec:internals-overview}
% --------------------------------------------------------------

Let's start to look at the internal having a general overview which is illustrated in 
Figure~\ref{fig:pipeline-overview}. This shows the stages that lead from a declared \py{Model} to a Feynman-rule \py{Expression}. We can summerize it in four layers, see (Table~\ref{tab:internals-layers}).

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=6mm,
    box/.style={draw, rounded corners, align=center, text width=8.4cm,
                minimum height=8mm, font=\small, inner sep=5pt},
    decl/.style={box, fill=blue!8},
    comp/.style={box, fill=orange!12},
    eng/.style={box, fill=green!12},
    post/.style={box, fill=purple!10},
    arr/.style={-{Latex[length=2.2mm]}, thick}
]

\node[decl]                      (model)     {\texttt{Model(...)}: declared fields, gauge groups, source Lagrangian};
\node[decl,  below=of model]     (declared)  {\texttt{DeclaredLagrangian.source\_terms} (still purely syntactic)};
\node[comp,  below=of declared]  (analyze)   {\texttt{\_analyze\_declared\_source\_term} $\to$ \texttt{AnalyzedSourceTerm} (routing)};
\node[comp,  below=of analyze]   (compile)   {\texttt{compiler.gauge}: expand \texttt{DC}, \texttt{FS}, gauge fixing, ghosts; re-lower};
\node[comp,  below=of compile]   (terms)     {tuple of \texttt{InteractionTerm} $\to$ \texttt{CompiledLagrangian}};
\node[eng,   below=of terms]     (legs)      {\texttt{feynman\_rule(...)}: build \texttt{ExternalLeg}s, \texttt{to\_vertex\_kwargs(...)}};
\node[eng,   below=of legs]      (contract)  {\texttt{contract\_to\_full\_expression}: sum over Wick contractions};
\node[post,  below=of contract]  (factor)    {\texttt{vertex\_factor}: overall $i$, optional momentum $\delta$, strip externals};
\node[post,  below=of factor]    (canon)     {idenso simplification + tensor canonicalisation};
\node[decl,  below=of canon]     (out)       {Symbolica \texttt{Expression}: the Feynman rule};

\foreach \a/\b in {model/declared, declared/analyze, analyze/compile,
                   compile/terms, terms/legs, legs/contract,
                   contract/factor, factor/canon, canon/out}
  \draw[arr] (\a) -- (\b);
\end{tikzpicture}
\caption{The pipeline from a declared \texttt{Model} to a Feynman-rule \texttt{Expression}. Blue = declaration layer (\texttt{feynpy}); orange = lowering/compiler layer (\texttt{feynpy.lowering}, \texttt{compiler.gauge} and helpers); green = contraction engine (\texttt{symbolic.vertex\_engine}); purple = post-processing (\texttt{symbolic.vertex\_postprocessing}, \texttt{symbolic.tensor\_canonicalization}).}
\label{fig:pipeline-overview}
\end{figure}

\begin{table}[htbp]
\centering
\begin{tabular}{@{}p{2.6cm}p{3.6cm}p{6.8cm}@{}}
\toprule
\textbf{Layer} & \textbf{Package(s)} & \textbf{Responsibility} \\
\midrule
Declaration &
\texttt{feynpy} (\texttt{metadata}, \texttt{declared}, \texttt{core}) &
Represents the physics as the user wrote it: declared fields, gauge groups, and source Lagrangian terms, still in a mostly syntactic form. \\
Lowering \& compilation &
\texttt{feynpy.lowering}, \texttt{lagrangian}, \texttt{compiler} &
Classifies and validates each source term, expands gauge structures such as covariant derivatives, field strengths, gauge-fixing terms and ghosts into local monomials, and collects the result into \texttt{InteractionTerm}s. \\
Contraction engine &
\texttt{symbolic.vertex\_engine} &
Builds external legs and vertex arguments, then sums over Wick contractions to assemble the full expression. \\
Post-processing &
\texttt{symbolic.vertex\_postprocessing}, \texttt{symbolic.tensor\_canonicalization} &
Applies the overall factor of $i$, optionally extracts the momentum-conserving $\delta$, strips external-leg factors, and simplifies and canonicalises the resulting tensor expression. \\
\bottomrule
\end{tabular}
\caption{The four layers of the pipeline, coloured as in Figure~\ref{fig:pipeline-overview}: declaration (blue), lowering/compilation (orange), contraction (green), and post-processing (purple).}
\label{tab:internals-layers}
\end{table}

The remaining sections walk the layers from the outside in: declaration (Section~\ref{sec:internals-declaration}), analysis and index resolution (Section~\ref{sec:internals-lowering}), gauge compilation (Section~\ref{sec:internals-compiler}), the \py{InteractionTerm} intermediate representation that ties everything together (Section~\ref{sec:internals-ir}), the Wick-contraction engine (Section~\ref{sec:internals-contraction}), and canonicalisation (Section~\ref{sec:internals-postprocessing}). Section~\ref{sec:internals-difficulty} then collects the hardest problems encountered while building the pipeline --- the places where the design had to be most deliberate.

\footnote{Each TikZ figure in this chapter has a plain-text counterpart kept as a Mermaid source diagram under \texttt{viz/mermaid/}}

% --------------------------------------------------------------
\section{The outer layer: declaring a model}
\label{sec:internals-declaration}
% --------------------------------------------------------------

\subsection{Inert metadata}

In the chapter \ref{ch:feynpy} we have see all the building blocks of FEYNPY that will enable us to define the Model. This building blocks are the declarative \emph{metadata} stored in \texttt{feynpy.metadata}, these are frozen dataclasses that hold physics data and carry no algorithmic logic. Let's summerize them here:
\begin{itemize}
    \item \py{IndexType} pairs a Spenso representation with a ``kind'' string and a label prefix (more in Section~\ref{sec:cas-indextype}); 
    \item \py{Field} stores a name, spin, self-conjugacy, its tuple of \py{indices}, its Symbolica symbol and conjugate symbol, quantum numbers, and optional flavour-class data, inferring its \py{kind} (scalar/fermion/vector) and \py{statistics} from the spin; 
    \item \py{Parameter} behaves algebraically like its symbol and may carry indices and explicit components; 
    \item \py{GaugeRepresentation} and 
    \item \py{GaugeGroup} record the symmetry data, whether a group is abelian, its coupling, gauge boson, ghost field, structure-constant builder and matter representations
\end{itemize}
    
The deliberate point is that metadata is inert: a faithful, machine-readable copy of what a physicist would write on paper, with all behaviour deferred downstream.
\cleanup{till here}
\subsection{The declarative DSL as wrapper objects}

Now that we have defined the building blocks of the theory, we can write the Lagrangian itself. This is done with a small domain-specific language (DSL) defined in \texttt{feynpy.declared}:\footnote{A domain-specific language is a small language designed for a particular task. Here it is a restricted set of Python objects and functions that lets the user write Lagrangian terms in a notation close to standard physics notation.} \py{DC} (covariant derivative), \py{FS} (field strength), \py{Gamma}, \py{PartialD}, \py{Metric}, \py{T}, \py{StructureConstant}, \py{GaugeFixing}, \py{GhostLagrangian}. The important point is that these constructors do not immediately build a Symbolica tensor expression. They build small immutable wrapper objects that will later be analysed and lowered into compiled interaction terms. When the user writes the first term of equation \ref{eq:qcd-full}
\begin{minted}{python}
I * Quark.bar * Gamma(mu) * DC(Quark, mu)
\end{minted}

Obviously the Lagrangian in equation \ref{eq:qcd-full} is composed of many terms, and so the first thing the code needs to do is to distinguish each monomial and separate its scalar coefficient from its field and tensor factors. More precisely the code assembles a \py{_DeclaredMonomial}. Its content is simply a scalar coefficient (\py{I}) and an ordered tuple that stores every factor --- here \py{_FieldFactor(Quark, conjugated=True)}, \py{GammaFactor(mu)}, \py{CovariantDerivativeFactor(Quark, mu)}. At this stage nothing is contracted, expanded or simplified. The order of the factors is preserved exactly because later passes need the original source structure. Later a process of analysis will determine what needs to be expanded.

A useful subtlety is that a few helpers are \emph{dual}. \py{Gamma(mu)} and \py{T(a)} are declarative placeholders: they say ``insert the gamma matrix'' or ``insert the generator'' and let the lowering pass infer the endpoints from the neighbouring fields. By contrast, \py{Gamma(i, j, mu)} and \py{T(a, i, j)} bypass that shorthand and return the fully explicit Spenso tensors \py{gamma(i, j, mu)} and \py{t(a, i, j)} immediately. This gives the user both a compact notation without specifying every index explicitly when the structure is obvious, and a fully explicit notation when it is not.

\subsection{\texttt{Model} and the two Lagrangian containers}

\py{Model} (in \texttt{feynpy.core}) sits one level above the objects of the previous subsection. There the basic unit was a single product term, represented by a \py{_DeclaredMonomial}. A full source Lagrangian is then just an ordered collection of such source terms, stored in the container \py{DeclaredLagrangian}. In other words, the hierarchy is:
\[
\text{factors} \;\longrightarrow\; \py{_DeclaredMonomial} \;\longrightarrow\; \py{DeclaredLagrangian} \;\longrightarrow\; \py{Model}.
\]
Strictly speaking, the entries of \py{DeclaredLagrangian.source\_terms} need not all be \py{_DeclaredMonomial}s: special declarations such as \py{GaugeFixing(...)} and \py{GhostLagrangian(...)} are stored there as source terms as well. But its role is always the same: it is the source-level container for the declared Lagrangian, and nothing more.

\py{Model} is the larger object that combines that declared Lagrangian with the rest of the model data: optional \py{fields}, \py{parameters} and \py{gauge_groups}. Its constructor accepts the declared Lagrangian either as the first positional argument or via \py{lagrangian_decl=}. The first step is normalization: whatever admissible source object the user passes is coerced into a \py{DeclaredLagrangian}. That container is deliberately minimal because its only payload is the ordered tuple \py{source_terms}; its job is to preserve the source-level declaration, not to compile it.

The second step is metadata completion. \py{Model} scans the source terms and collects the fields and gauge groups that appear in them: ordinary field factors, \py{DC(...)} and \py{PartialD(...)} factors, field strengths, kinetic declarations, gauge-fixing declarations and ghost declarations all contribute information. From the discovered gauge groups it also adds the associated gauge boson and ghost fields, and from each field it also includes any declared flavour-class members. This is why simple models often need very little explicit boilerplate.

The third step is early validation. For every source term, \py{Model} calls the analyser of Section~\ref{sec:internals-lowering}, namely \py{_analyze_declared_source_term(...)}, and checks that the term can be classified and routed through the compiler. If a monomial is not a supported local interaction, covariant core, field-strength term, gauge-fixing term, ghost term or generic covariant monomial, model construction fails immediately. In other words, unsupported syntax is rejected when the model is declared, not only later when one asks for a Feynman rule.

Actual compilation is lazy and cached. The first call to \py{model.lagrangian()} invokes \texttt{compiler.gauge}, builds a \py{CompiledLagrangian}, and stores it for reuse. The distinction between \py{DeclaredLagrangian} and \py{Model} is therefore simple: \py{DeclaredLagrangian} is just the source-level Lagrangian by itself, whereas \py{Model} is the full object that owns that Lagrangian together with its metadata and the compilation machinery. The distinction between the two Lagrangian containers is also strict by design: \py{DeclaredLagrangian} stores source-level objects, while \py{CompiledLagrangian} stores lowered \py{InteractionTerm} objects and is the entry point for later stages such as flavour expansion, Symbolica export and vertex extraction. Their algebra is intentionally separated: declarative objects add only to declarative objects, compiled objects only to compiled objects, and mixing the two raises an explicit error that sends the user back through \py{model.lagrangian()}. This keeps the boundary between the ``source'' and ``compiled'' representations precise.

% --------------------------------------------------------------
\section{Analysis and lowering: reading the Lagrangian and resolving indices}
\label{sec:internals-lowering}
% --------------------------------------------------------------

The lowering layer (\texttt{feynpy.lowering}, with model-agnostic helpers in \texttt{lagrangian.lowering}) is the bridge between declarative syntax and the compiler. For each source term it answers three questions: what kind of term is this, are its indices well-formed, and --- if the term is already local --- what is its canonical interaction form?

\subsection{Classifying each source term}

The single entry point is \py{_analyze_declared_source_term(...)}, which returns an \py{AnalyzedSourceTerm}. This record is the model layer's source of truth: the compiler consumes it and never re-derives the classification. The categories are summarised in Table~\ref{tab:analyzed-categories}.

\begin{table}[htbp]
\centering
\begin{tabular}{@{}p{4.0cm}p{5.0cm}p{3.2cm}@{}}
\toprule
\textbf{Category} & \textbf{Meaning} & \textbf{Example} \\
\midrule
\texttt{interaction} & already a local \py{InteractionTerm}; no compilation & \py{lam * phi**4} \\
\texttt{covariant\_core} & canonical kinetic $\bar\psi\, i\gamma^\mu D_\mu \psi$ or $(D_\mu\Phi)^\dagger(D^\mu\Phi)$ & quark kinetic term \\
\texttt{field\_strength\_monomial} & a product of \py{FS(...)} factors & $-\tfrac14 F^a F^a$ \\
\texttt{generic\_covariant\_monomial} & any other monomial containing \py{DC} & current$\,\times\,$current \\
\texttt{gauge\_fixing} & a gauge-fixing declaration & \py{GaugeFixing(SU3, xi)} \\
\texttt{ghost} & a ghost declaration & \py{GhostLagrangian(SU3)} \\
\bottomrule
\end{tabular}
\caption{The classification produced by \texttt{\_analyze\_declared\_source\_term}. Classification is priority-ordered (top to bottom), and every category except \texttt{interaction} has \texttt{needs\_compilation = True}.}
\label{tab:analyzed-categories}
\end{table}

Classification is priority-ordered: the analyser tries ``already local'' first, then the canonical covariant kinetic shapes, then field strengths, then gauge fixing, then ghosts, and finally the general covariant-monomial catch-all.

\subsection{Resolving the indices}

This is where ``expanding the indices'' actually happens, and it received a disproportionate share of the design effort because it is where silent errors are most dangerous.

\paragraph{Monomial-wide label validation.} In the compact notation the user writes plain labels such as \py{mu}, \py{a}, \py{c}, \py{f}. On paper these are unambiguous because a human reads the whole term at once; the code must reproduce that global reading. Before anything else, the analyser builds one registry mapping every label \emph{name} used anywhere in a monomial --- on field slots, on declared factors like \py{PartialD}, \py{Gamma}, \py{T}, $f^{abc}$, \py{FS}, and even on raw Spenso tensors in the coefficient --- to exactly one \py{IndexType}. If a name is used for two incompatible index kinds (say a colour label reused as a generation label), the term is rejected with an explicit error. This closes a real hole: earlier the labels were validated only slot-locally, so an inconsistent term like \py{q.bar(f, c) * q(c, f) * Phi} could compile into a wrong contraction.

\paragraph{Slot binding and inference.} Once labels are typed, local monomials pass through a small state machine that, in a fixed order: (1)~seeds any explicit labels the user wrote; (2)~applies chain factors (\py{Gamma}, \py{T}) between adjacent fields, connecting spinor and colour endpoints; (3)~binds remaining typed labels to open field slots \emph{by kind}, requiring a one-to-one match and raising an ambiguity error otherwise; (4)~shares fresh labels across natural pairs (an adjacent $\bar\psi/\psi$, a unique open bar/field pair); and (5)~invents fresh dummy labels, using the index-type prefix, for whatever remains. The result is that a physicist can write very little and still obtain a fully and unambiguously indexed term --- or a clear error when what they wrote cannot be resolved uniquely.

\subsection{Remembering fermion pairings}

When a term is already local, lowering builds an \py{InteractionTerm} directly (Section~\ref{sec:internals-ir}) and, during the same step, records which fermion occurrences form closed Dirac bilinears --- the \py{closed_dirac_bilinears} metadata: a list of $(\bar\psi\text{-slot},\,\psi\text{-slot})$ pairs, inferred from gamma-matrix chains, from repeated spinor labels, or from the spinor graph of the coefficient. It records the physicist's intent (which current pairs with which) so the contraction engine can later apply the correct fermion sign. Its importance is discussed in Section~\ref{sec:internals-difficulty}.

% --------------------------------------------------------------
\section{Compilation: turning gauge structure into ordinary interactions}
\label{sec:internals-compiler}
% --------------------------------------------------------------

The compiler (\texttt{compiler.gauge} and its helpers) is where the physics of gauge symmetry is applied. Its top-level function \py{_compile_declared_lagrangian_terms(model)} walks the source terms and dispatches each analysed term to a specialised handler (Figure~\ref{fig:compiler-dispatch}). The governing conventions are fixed once, in the module:
\begin{equation}
    D_\mu = \partial_\mu - i g A_\mu,
    \qquad
    F^a_{\mu\nu} = \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g f^{abc} A^b_\mu A^c_\nu .
    \label{eq:conventions}
\end{equation}

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=5mm and 9mm,
    box/.style={draw, rounded corners, align=center, font=\footnotesize,
                inner sep=4pt, minimum height=7mm},
    comp/.style={box, fill=orange!12, text width=3.9cm},
    hand/.style={box, fill=orange!6, text width=4.6cm},
    ir/.style={box, fill=green!12, text width=2.6cm},
    arr/.style={-{Latex[length=2mm]}, thick}
]
\node[comp] (c1) {covariant\_core};
\node[comp, below=of c1] (c2) {generic\_covariant\_monomial};
\node[comp, below=of c2] (c3) {field\_strength\_monomial};
\node[comp, below=of c3] (c4) {gauge\_fixing};
\node[comp, below=of c4] (c5) {ghost};

\node[hand, right=of c1] (h1) {\texttt{\_compile\_declared\_covariant\_core}};
\node[hand, right=of c2] (h2) {\texttt{\_compile\_generic\_declared\_covariant\_monomial}};
\node[hand, right=of c3] (h3) {\texttt{compile\_field\_strength\_monomial}};
\node[hand, right=of c4] (h4) {\texttt{compile\_gauge\_fixing\_term}};
\node[hand, right=of c5] (h5) {\texttt{compile\_ghost\_term}};

\node[ir, right=of h3] (ir) {\texttt{InteractionTerm} tuple};

\foreach \c/\h in {c1/h1, c2/h2, c3/h3, c4/h4, c5/h5} \draw[arr] (\c) -- (\h);
\foreach \h in {h1, h2, h3, h4, h5} \draw[arr] (\h) -- (ir);
\end{tikzpicture}
\caption{Compiler dispatch. Each analysed category is handled by one routine; every routine ultimately returns ordinary \texttt{InteractionTerm} objects. The structured categories are \emph{expanded} into local monomials and re-lowered through the same path as hand-written interactions.}
\label{fig:compiler-dispatch}
\end{figure}

\subsection{The expand-then-lower principle}

The single most important decision in the compiler is that it \emph{does not build vertices from gauge operators directly}. Instead, each structured operator is expanded into ordinary local monomials --- partial-derivative pieces, gauge-current pieces, and explicit generator or structure-constant factors --- and those monomials are sent back through the same local lowering used for hand-written interactions (Section~\ref{sec:internals-lowering}). The payoff is large: there is no separate compiler for $F^2$, $F^3$ or $F^4$, and none for each covariant power; the combinatorics come for free from distribution plus one well-tested lowering path. The cost is that raw compiled output is more expanded than a textbook form --- a non-abelian $F^2$ becomes nine local monomials rather than one --- which is later collapsed by canonicalisation (Section~\ref{sec:internals-postprocessing}).

\subsection{Covariant derivatives}
\label{sec:internals-covd}

Structurally, a covariant derivative is ``a partial derivative plus one gauge current for every group acting on the field''. With the convention of Eq.~\eqref{eq:conventions}, the general form the compiler implements is
\begin{equation}
    D_\mu \phi \;=\; \partial_\mu \phi \;-\; i \sum_{G}\, g_G\, A^{(G)}_\mu\,\rho_G(\phi),
    \label{eq:covd-general}
\end{equation}
where the sum runs over every declared gauge group $G$ under which $\phi$ transforms, $g_G$ is that group's coupling, $A^{(G)}_\mu$ its gauge boson, and $\rho_G(\phi)$ the action of the group on the representation index that $\phi$ carries. The two cases differ only in what $\rho_G$ is:
\begin{itemize}
    \item \textbf{abelian} --- $\rho_G(\phi) = Q_G\,\phi$, with $Q_G$ the field's charge read from its quantum numbers, so the piece is $-\,i\,g_G\,Q_G\,A^{(G)}_\mu\,\phi$ (no generator, no adjoint index);
    \item \textbf{non-abelian} --- $\rho_G(\phi)$ acts with a generator in the field's representation, so the piece is $-\,i\,g_G\,A^{a}_{G,\mu}\,(t^a_G)_{ij}\,\phi_j$, and the gauge boson carries the adjoint label $a$.
\end{itemize}

The routine \py{expand_cov_der} realises Eq.~\eqref{eq:covd-general} literally: it resolves the set of gauge groups the field transforms under --- and if the user named none, it \emph{sums over all of them} --- and emits one gauge piece per group plus a single partial-derivative piece. A conjugated occurrence $(D_\mu\phi)^\dagger$ flips the sign of every gauge piece (the coupling picks up $+i$ instead of $-i$).

\paragraph{Several slots of the same representation.} If the field carries more than one index in the acting representation, the generator must hit each slot in turn, and Eq.~\eqref{eq:covd-general} expands into a sum over slots. For a field $\phi_{i_1 i_2}$ with two fundamental colour indices,
\begin{equation}
    D_\mu \phi_{i_1 i_2}
    = \partial_\mu \phi_{i_1 i_2}
    - i g\, A^a_\mu\Big[(t^a)_{i_1 j}\,\phi_{j\, i_2} + (t^a)_{i_2 j}\,\phi_{i_1 j}\Big].
    \label{eq:covd-multislot}
\end{equation}
Whether this multi-slot expansion is performed at all is controlled by the representation's \py{slot_policy}: the strict default \emph{rejects} an ambiguous repeated-index field, and \py{slot_policy='sum'} opts in to the sum of Eq.~\eqref{eq:covd-multislot} (with one generator insertion per active slot).

\paragraph{Several gauge groups at once.} The prototypical case is a field transforming under three groups simultaneously. For the left-handed quark doublet $Q_L$ of the Standard Model, charged under $\mathrm{SU}(3)_c\times\mathrm{SU}(2)_L\times\mathrm{U}(1)_Y$, Eq.~\eqref{eq:covd-general} reads
\begin{equation}
    D_\mu Q_L
    = \partial_\mu Q_L
    \;-\; i g_s\, G^a_\mu\, t^a_{(3)}\,Q_L
    \;-\; i g\, W^I_\mu\, t^I_{(2)}\,Q_L
    \;-\; i g'\, Y_{Q}\, B_\mu\, Q_L ,
    \label{eq:covd-multigroup}
\end{equation}
with $t^a_{(3)}$ the colour-fundamental generators, $t^I_{(2)}$ the weak-isospin generators and $Y_Q$ the hypercharge. Writing \py{DC(Q_L, mu)} with no group named produces exactly these four pieces --- the plain $\partial_\mu Q_L$ plus one gauge current per group; naming a single group, \py{DC(Q_L, mu, gauge_group=SU3C)}, keeps only that group's current together with the derivative.

For the canonical fermion kinetic term of Eq.~\eqref{eq:qcd-current} the covariant-core handler builds the free part $i\bar\psi\gamma^\mu\partial_\mu\psi$ and one fermion-current interaction per gauge piece --- for QCD the $\bar q\,\gamma^\mu t^a q\,G^a_\mu$ vertex with coupling $-ig_s$. For a complex scalar $(D_\mu\Phi)^\dagger(D^\mu\Phi)$ the same machinery produces both the single-boson current and the two-boson contact (``seagull'') term, and, when the scalar transforms under two groups at once, the mixed cross-group contact terms $\propto g_r g_s\,A^{(r)}_\mu A^{(s)\mu}$ as well.

\subsection{Field strengths}
\label{sec:internals-fs}

The gauge-invariant kinetic term is built from the field strength, which with the same convention is
\begin{equation}
    F^a_{\mu\nu} = \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g\, f^{abc}\, A^b_\mu A^c_\nu ,
    \label{eq:fs-general}
\end{equation}
degenerating for an abelian group to $F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu$ (no adjoint index, no cubic term). This is written in the DSL as \py{FS(SU3, mu, nu, a)} in the non-abelian case --- the trailing $a$ is the adjoint index --- and as \py{FS(U1, mu, nu)} in the abelian case, which must omit it.

A single \py{FS} factor is expanded into its additive pieces: for a non-abelian group the three pieces of Eq.~\eqref{eq:fs-general},
\begin{equation}
    P_1 = \partial_\mu A^a_\nu, \qquad
    P_2 = -\,\partial_\nu A^a_\mu, \qquad
    P_3 = g\, f^{abc}\, A^b_\mu A^c_\nu ,
    \label{eq:fs-pieces}
\end{equation}
and only $P_1, P_2$ for an abelian group. A monomial containing several \py{FS} factors is then expanded by distributing the \emph{full Cartesian product} of these pieces. Take pure Yang--Mills for $\mathrm{SU}(3)$,
\begin{equation}
    \mathcal{L}_{\mathrm{YM}} = -\tfrac14\, F^a_{\mu\nu} F^{a\,\mu\nu},
\end{equation}
where each of the two field strengths contributes the three pieces of Eq.~\eqref{eq:fs-pieces}. The product distributes into the full $3\times 3 = 9$ local monomials, which fall into three physical sectors:
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Piece combination} & \textbf{Count} & \textbf{Sector} \\
\midrule
$(P_1,P_2)\times(P_1,P_2)$ & $2\times 2 = 4$ & two-gluon (kinetic) \\
$(P_1,P_2)\times P_3$ and $P_3\times(P_1,P_2)$ & $2 + 2 = 4$ & three-gluon vertex \\
$P_3\times P_3$ & $1$ & four-gluon vertex \\
\bottomrule
\end{tabular}
\end{center}
Each of the nine monomials is then lowered locally through the same path as a hand-written interaction. Because the mechanism is just distribution, higher operators need no new code: $f\,F^3$ yields $3^3 = 27$ monomials, $F^4$ yields $3^4 = 81$, and mixed-group products such as $F_{\mathrm{SU}(3)}^2\,F_{\mathrm{SU}(2)}^2$ are simply larger Cartesian products over the pieces of each factor. A scalarity check enforces that every adjoint label appears an even number of times, i.e.\ that the operator is a genuine gauge singlet.

\subsection{Gauge fixing and ghosts}

The gauge-fixing helper compiles $-\tfrac{1}{2\xi}(\partial\!\cdot\!A)^2$ by building the corresponding local derivative monomial with the right metric contractions, lowering it, and tagging its provenance as \py{gauge_fixing} so it can be recognised later by the reporting and validation layers; manually written $(\partial\!\cdot\!A)^2$ terms are detected and tagged identically, so helper and hand-written forms behave the same. The ghost handler compiles the non-abelian Faddeev--Popov sector $-\bar c\,\partial^\mu D_\mu c$ into its two local pieces: the bilinear $(\partial_\mu \bar c)(\partial^\mu c)$ with the adjoint metric, and the ghost--gauge vertex $g f^{\bar a a b}(\partial_\mu \bar c) A^\mu c$ with the group's structure constant. Because a ghost transforms in the adjoint, its covariant derivative reuses the same representation-aware machinery as matter fields rather than a ghost-only code path.

% --------------------------------------------------------------
\section{The pivot: the \texttt{InteractionTerm} intermediate representation}
\label{sec:internals-ir}
% --------------------------------------------------------------

Every branch above converges on one data type, \py{InteractionTerm} (in \texttt{feynpy.interactions}). It is the waist of the hourglass: everything above it is declarative and physics-aware, everything below it is generic symbolic machinery.

\begin{minted}{python}
@dataclass(frozen=True)
class InteractionTerm:
    coupling: object                        # Symbolica Expression
    fields: tuple[FieldOccurrence, ...]     # bare field factors, in order
    derivatives: tuple[DerivativeAction, ...] = ()
    closed_dirac_bilinears: tuple[tuple[int, int], ...] = ()
    label: str = ""
    sector: str = ""
\end{minted}

The \py{coupling} is a Symbolica expression holding the numerical and tensor prefactor (couplings, generators, structure constants, metrics); \py{fields} is an ordered tuple of \py{FieldOccurrence} objects, each a field with its slot-resolved index labels and conjugation flag; \py{derivatives} records, as \py{DerivativeAction(target, lorentz_index)}, which field each derivative hits and along which Lorentz index; and \py{closed_dirac_bilinears} carries the fermion-pairing provenance of Section~\ref{sec:internals-lowering}. A \py{CompiledLagrangian} is simply a tuple of such terms. Note what has \emph{not} happened yet: no creation operators, no permutations, no external momenta. An \py{InteractionTerm} is a purely static, index-labelled description of one monomial of the Lagrangian --- the direct code counterpart of the right-hand side $g_{\alpha_1\cdots\alpha_n}\,\partial\cdots\partial\,\phi^{\alpha_1}\cdots\phi^{\alpha_n}$ of Eq.~\eqref{eq:general-term}.

The ordering of \py{fields} is significant. Because Symbolica multiplication is commutative, the ordered tuple --- not the exported Symbolica expression --- is the authoritative record of fermion order; this is why the whole pipeline carries fermion structure as explicit metadata rather than trusting the algebra to remember it. The method \py{to_vertex_kwargs(external_legs)} performs the last model-aware step: it strips away all model-layer vocabulary and reduces one term plus a chosen set of external legs to flat, parallel lists --- species keys, momenta, field and leg roles, index labels and types, derivative indices and targets, and the bilinear metadata --- after which only Symbolica and Spenso objects remain.

\subsection*{Flavour expansion}

If a field carries a flavour index (Section~\ref{sec:feynpy-flavor}), terms stay generation-generic by default: a single term $\lambda_{ij}\,\bar\ell_i \ell_j\,\phi$ implicitly sums over $i,j$. Requesting \py{flavor_expand=True} instead enumerates every member combination, substitutes the corresponding \py{Parameter} component (e.g.\ $\lambda_{e\mu}$), and \emph{drops} the term when that component was declared zero, producing one fully explicit \py{InteractionTerm} per combination in the same data structure.

% --------------------------------------------------------------
\section{The inner core: Wick contraction in the vertex engine}
\label{sec:internals-contraction}
% --------------------------------------------------------------

The vertex engine (\texttt{symbolic.vertex\_engine}) is deliberately ignorant of physics vocabulary. It sees fields, roles, index labels, momenta and a coupling tensor, and performs a combinatorial contraction that is the code-level realisation of the canonical-quantisation algorithm of Section~\ref{sec:feynrules-algorithm}.

\subsection{The permutation sum}

The core routine \py{contract_to_full_expression} sums over all $n!$ ways of matching the $n$ external legs to the $n$ field slots of the interaction. Each permutation \py{perm} assigns external leg \py{perm[i]} to field slot \py{i}, and the sum over all of them is the full Wick contraction. For each surviving permutation the code builds one contribution:

\begin{enumerate}
    \item \textbf{Coupling remap.} Open index labels in the coupling (those appearing exactly once across all slots) are relabelled to the labels of the external legs they now contract with --- literally \py{coupling.replace(open_label, target_label)}, using the Symbolica pattern machinery of Section~\ref{sec:cas-symbolica}.
    \item \textbf{Compatibility pruning.} Each $(\text{slot},\text{leg})$ pairing is checked by \py{factor_leg_compatible}. The check compares stable \emph{match keys} built from field identity (name, spin, kind, statistics, index schema, conjugation), which lets aliased external species still match the right field, and then checks role compatibility. An incompatible permutation is discarded immediately, before any further work.
    \item \textbf{Fermion sign.} For a fermionic term a Grassmann sign is applied from the permutation restricted to the fermion slots --- unless the closed-bilinear exception applies (Section~\ref{sec:internals-contraction-fermions}).
    \item \textbf{Derivatives $\to$ momenta.} Each \py{DerivativeAction} contributes a factor $(-i)\,p^\mu$, where $p$ is the momentum of the leg matched to the derivative's target. This realises the Fourier convention $\partial_\mu \to -i p_\mu$ with all momenta incoming.
    \item \textbf{External factors.} Each leg contributes its external wavefunction ($U$ for bosons, $UF/\overline{UF}$ for fermions) together with a species Kronecker delta, and repeated explicit indices are contracted through the appropriate Spenso representation metric.
    \item \textbf{Plane wave.} The contribution is multiplied by $\exp(-i\,\Sigma p\!\cdot\!x)$ and added to the running total.
\end{enumerate}

For a purely bosonic vertex all signs are $+1$, so the sum is a permanent over compatible pairings; for fermions each term carries its sign. The essential shape of the routine is:

\begin{minted}{python}
for perm in permutations(range(n)):
    term = coupling_with_open_indices_remapped_along(perm)
    if not all(factor_leg_compatible(i, perm[i], ...) for i in range(n)):
        continue                       # prune impossible pairing
    if statistics == "fermion" and not use_closed_dirac_bilinear_signs:
        term *= fermion_sign(perm, fermion_slots)
    for mu, tgt in zip(derivative_indices, derivative_targets):
        term *= (-I) * pcomp(ps[perm[tgt]], mu)
    for i, j in enumerate(perm):
        term *= external_factor(role[i], alphas[i], betas[j], ps[j], ...)
    term *= plane_wave(sum_of_matched_momenta, x)
    total += term
\end{minted}

\begin{tcolorbox}[infobox, title={Why this generalises so well}]
Nothing in \py{contract_to_full_expression} is specific to Lorentz or colour indices: an index is just a \py{(kind, label)} pair, and ``open'' versus ``internal'' is determined purely by how many times a label occurs. This is why the same engine derives QED, QCD, electroweak and ghost vertices without a single index-kind-specific branch.
\end{tcolorbox}

\subsection{Fermion signs and the closed-bilinear exception}
\label{sec:internals-contraction-fermions}

Fermion signs were the single most delicate piece of physics in the engine. The generic rule is the flat Grassmann sign of the fermion permutation. But this gives the \emph{wrong relative sign} for a product of closed Dirac bilinears compared with the FeynRules convention: for identical-fermion current--current operators the flat rule produces $\text{direct} - \text{crossed}$, whereas the reference gives $\text{direct} + \text{crossed}$. The resolution is metadata-driven and narrow: when --- and only when --- every Dirac fermion slot is covered exactly once by the disjoint \py{closed_dirac_bilinears} pairs carried down from lowering, the engine suppresses the flat permutation sign and uses the all-plus bilinear convention. Generic open multi-fermion tensors keep the flat Grassmann sign. This is exactly the right boundary: it adopts the convention where the term genuinely is a product of closed currents, and nowhere else.

\subsection{Output policy and universal factors}

The public wrapper \py{vertex_factor} calls the contraction and then applies an \emph{output policy}. With \py{include_delta=True} the accumulated plane wave $\exp(-i\,\Sigma p\!\cdot\!x)$ is replaced by the momentum-conservation factor $(2\pi)^d\,\delta(\Sigma p)$ (the high-level API omits this by default). With \py{strip_externals=True} (the default) the external wavefunctions $U$, $UF$, $\overline{UF}$ are amputated, giving the bare vertex. Finally the whole result is multiplied by the universal overall $+i$ that every vertex carries. The separation is intentional: the contraction is pure combinatorics, while the choice of whether to display the $\delta$ or amputate the legs is a presentation policy layered on top.

% --------------------------------------------------------------
\section{Making the answer canonical}
\label{sec:internals-postprocessing}
% --------------------------------------------------------------

The raw contracted expression is correct but not unique: the same vertex can be written with different dummy-index names, different orderings of antisymmetric tensors, or in an over-expanded field-strength basis. To compare against a reference (or against itself across runs), it must be brought to a canonical form. Two passes do this.

First, \texttt{symbolic.vertex\_postprocessing} applies the model-aware simplifications: it contracts species Kronecker deltas, collapses repeated bispinor indices via the Spenso metric, and, on request, runs the full Dirac algebra on closed gamma chains through idenso (Section~\ref{sec:cas-idenso}).

Second, \texttt{symbolic.tensor\_canonicalization} puts the surviving expression into canonical form on top of Symbolica's \py{canonize_tensors} (Section~\ref{sec:cas-symbolica}). Because that primitive understands symmetry only through the \py{is_symmetric}/\py{is_antisymmetric} flags on a plain symbol --- not through Spenso tensor heads --- the module temporarily \emph{swaps} every Spenso tensor (metric, generator, structure constant, $\gamma$, and the project's custom invariants) for a plain symbol carrying the correct symmetry, canonicalises, and swaps back. This yields deterministic dummy-index names without losing the typed Spenso representation used everywhere else. Index groups of different kinds (Lorentz, colour, spinor, weak) are kept separate so a dummy of one kind is never identified with a dummy of another. On top of this generic pass, \py{canonize_full} adds a few physics-specific reductions used when checking identities (gauge variation, BRST, Sections~\ref{sec:feynpy-gauge-variation} and~\ref{sec:feynpy-brst}): commuting flat partial derivatives so $\partial_\mu\partial_\nu$ and $\partial_\nu\partial_\mu$ agree, reducing repeated structure constants with the Jacobi identity while staying in the $f$-basis, and dropping Yang--Mills terms that vanish by the antisymmetry of $f^{abc}$ after a legal relabelling of dummy indices.

This layered canonicalisation is what turns the deliberately over-expanded compiler output back into a compact, comparable Feynman rule. Every intermediate object along the way --- \py{InteractionTerm}, \py{ExternalLeg}, the flat contraction inputs, the raw contraction, the canonical result --- remains fully inspectable, which is what makes the toolkit usable both as a Feynman-rule calculator and as a general symbolic Lagrangian-manipulation environment (Section~\ref{sec:feynpy-apply-operator} onward).

% --------------------------------------------------------------
\section{Where the difficulty lived}
\label{sec:internals-difficulty}
% --------------------------------------------------------------

Building the pipeline was less about any single algorithm than about a handful of recurring hard problems. These are the places where the most design attention was spent, and they are worth naming explicitly because they shaped the architecture.

\paragraph{Index bookkeeping and repeated slots.} The hardest bug class came from indices, not algebra. Early code keyed labels by index \emph{kind}, which collapses the moment a field carries two slots of the same kind (two Lorentz indices, two colour indices). The fix runs deep: slot-ordinal-stable label packing on fields; monomial-wide typed-label validation binding each label name to one index type; a deterministic multi-phase slot binder that fails loudly on ambiguity rather than guessing; and an explicit \py{slot}/\py{slot_policy} choice on gauge representations. The guiding rule became \emph{fail fast}: an ambiguous index assignment is an error at declaration time, never a silent wrong contraction.

\paragraph{Fermion signs.} Matching the FeynRules sign convention for products of closed Dirac bilinears without breaking generic multi-fermion terms (Section~\ref{sec:internals-contraction-fermions}) required threading explicit bilinear provenance from lowering, through the compiler and spectator decoration, into the contraction engine. Getting the boundary exactly right --- all-plus only when every fermion slot is covered by disjoint closed pairs --- took careful tracing on benchmark operators and is now frozen as a documented convention.

\paragraph{Conventions.} Signs and normalisations for $D_\mu$, $F^a_{\mu\nu}$, the Fourier factor $\partial_\mu\to -ip_\mu$, the universal $+i$, gauge fixing and ghosts were a persistent source of spurious ``bugs'' until they were frozen in one reference (Eq.~\eqref{eq:conventions}) and enforced by tests. Separating the structural compiler from the convention-fixed physical compiler removed a whole family of sign confusions.

\paragraph{Making expansion generic.} Replacing the special-case $F^2$ recogniser with a generic \py{FS} expansion --- and, symmetrically, the expand-then-lower covariant path --- unlocked $F^3$, $F^4$ and mixed-group operators for free. But it also multiplied the number of raw local monomials and exposed a performance cliff in the contraction engine, because the old compatibility check compared only broad field \emph{roles} and therefore iterated over impossible cross-species pairings. The species-aware match-key pruning of Section~\ref{sec:internals-contraction} was the direct response, restoring interactive performance on large mixed vertices without changing a single resulting formula.

\paragraph{Keeping the layers honest.} Finally, much of the effort went into \emph{not} letting the layers leak: declarative syntax must never reach the contraction engine, the engine must never learn what a covariant derivative is, and comparison against external references must live outside the engine entirely. This discipline --- one job per layer, one intermediate representation at the waist --- is what ultimately made it possible to validate the packaged Standard Model against the reference export at the level of every tree-level interaction vertex.

% --------------------------------------------------------------
\section{Summary}
\label{sec:internals-summary}
% --------------------------------------------------------------

The pipeline is a chain of narrow, inspectable rewrites. A \py{Model} holds inert metadata and a source-form Lagrangian. Lowering classifies each term, validates its indices globally, and lowers local terms to \py{InteractionTerm}s. The compiler expands gauge structure --- covariant derivatives, field strengths, gauge fixing and ghosts --- into ordinary local monomials and re-lowers them, so all the gauge physics funnels through one path. Every branch produces the same \py{InteractionTerm} intermediate representation, which the symbolic engine contracts by a pruned permutation sum, applying momentum factors, external wavefunctions, fermion signs and the universal $+i$, before canonicalisation makes the result unique. The recurring difficulties --- index bookkeeping, fermion signs, conventions, generic expansion and layer discipline --- are precisely the places where the design had to be most deliberate, and they are what the validation of the following chapter ultimately rests on.
